Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2024-01-18 analysis/paper/paper.Rmd
Updated figure formating for JAMT

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Department of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: true
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  # \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  By developing a new methodology for handling and assessing a large number of shoreline dated sites, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway. Both measures have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating. The findings indicate a largely diverging development of the two data sets through the Mesolithic. The number of shoreline dated sites undergoes some process of overall decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period. The precise nature of this discrepancy will necessitate further substantiation and the method of shoreline dating itself still requires further testing and assessment. Having noted this, we still tentatively suggest that while the number of shoreline dated sites is not devoid of influence from a demographic signal, this measure could be more heavily influenced by mobility patterns. Conversely, we also suggest that the lacking radiocarbon data from the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data more directly reflects population dynamics.
keywords: | 
  Mesolithic; Shoreline dates; Radiocarbon dates; Fennoscandia
---
Keywords: `r rmarkdown::metadata$keywords`

@@

Previously, the frequency of shoreline dated sites has been evaluated by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g. @fossum2020; @solheim2018; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021]. However, the point-based estimates do not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the north-east to Arendal municipality in the south-west (see map in Figure \@ref(fig:map)). The result of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.

@@

Summing the probability of the shoreline dated sites and the model-fitting procedures followed the same structure as that for radiocarbon dates and was partly based on re-purposed programming code from *rcarbon* and *ADMUR*. However, idiosyncrasies in the dating method did necessitate some adjustments. To illustrate this, the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Figure \@ref(fig:shoredate) and outlined below.

Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B). These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory of relative sea-level (RSL) change has been the same [see @svendsen1987, and map in Figure \@ref(fig:shoredate)A]. The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated. This is done by inverse distance weighting [e.g. @conolly2020], where the RSL change is interpolated to the location of an archaeological site based on its distance from the isobases of the geological displacement curves.

@@

With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline. Figure \@ref(fig:shoredate)C indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve. As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date. However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken. Furthermore, the common practice of adding a constant error or binning point-estimates of shoreline dates, as mentioned above, does not consider that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.

@@

The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833. This is incorporated in the procedure for shoreline dating used for this study (Figure \@ref(fig:shoredate)D). When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure thus gives the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore be commutative [see @weninger2015, 545], each shoreline date is normalised to sum to unity before being summed. To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.

@@

The implications of different approaches to shoreline dating and summing the results are illustrated in Figure \@ref(fig:sumshore). In Figure \@ref(fig:sumshore)A the sites to be shoreline dated are plotted according to their elevation. Due to the variable rates of shoreline displacement within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly inform us about their temporal distribution. For example, as the trajectory for sea-level regression is more pronounced towards the north-east, the same elevation at a location to the south-west implies a younger date compared to one situated further north-east.

When the frequency distribution of shoreline dates has been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking the mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C). As outlined above,  these point-estimates are then typically aggregated in bins of 200 or 500 years. This approach is illustrated in Figure \@ref(fig:sumshore)B. The SSPD resulting from using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is indicated by the shifts in the positions of peaks and troughs in the distribution after c. 8000 BCE.

@@

In the same way as with an RSPD, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above. Furthermore, similar to how characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of RSL change effectively functions as the calibration curve for each site. As the shoreline displacement curves are interpolated to the sites based on their location along a south-west/north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on exactly the same isobase. With analogy to the radiocarbon methodology, this would be equivalent to each radiocarbon date being associated with a unique calibration curve. As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2 km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites. These intervals were assigned a weight based on how the density of observed sites is distributed among them (Figure \@ref(fig:incpolys)).

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline-dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
The results of the Monte Carlo simulations are given in Figure \@ref(fig:spdmc), displaying the deviations from the simulation envelopes and the corresponding global p-values. The comparison of the models based on BIC scores is given in Figure \@ref(fig:bic).

For the RSPD, the logistic model achieved a p-value of 0.072 and could therefore not be rejected (Figure \@ref(fig:spdmc)E). As is also indicated by the BIC values of the compared models, the logistic model represents the best explanation of the data, and the failure to reject the model means that the data could be expected under this logistic development. This is characterised by a rapid growth from the start of the RSPD, around 8500 BCE, which plateaus at around 7500 BCE and then remains stable for the remainder of the period.

All the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline-dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A). Some significant deviations occur with the model over-predicting the frequency of sites in the period from c. 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE, lasting until around 8500 BCE. The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation between c. 6800 and 5800 BCE. Finally, a positive deviation occurs again around 5300--5000 BCE. While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [@roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded in the discussion to follow.

@@

Another relevant point to mention here is that the RSL change is effectively built into the Monte Carlo simulations, much in the same way as the Monte Carlo simulation method used with radiocarbon dates account for fluctuations in the calibration curve [e.g. @crema2022]. The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Figure \@ref(fig:incpolys)A), the corresponding local trajectories of RSL change (Figure \@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma function employed for the dating procedure (Figure \@ref(fig:shoredate)D). Beyond uncertainties to do with the dating method itself, and by extension also the generation of the Monte Carlo envelope, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models. An increased movement instantiated by increased rates of RSL change could result in the transition to a settlement pattern involving more frequent moves than that which could be expected under a constant adjustment to the shoreline. If this is the case, the proxy would still reflect the degree of occupational duration, but this could be a response to the rate of RSL change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments. Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the substantive implications of the fluctuations in this proxy.

@@

# Acknowledgement{-}
We would like to thank David Wright for commenting on an earlier draft of this paper. We would also like to thank the reviewers for their constructive feedback and PCI archaeology for facilitating their evaluation of our manuscript.

>>>>

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Department of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: true
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  # \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/apa.csl" # Insert path for the bib-style
abstract: |
  By developing a new methodology for handling and assessing a large number of shoreline dated sites, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway. Both measures have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating. The findings indicate a largely diverging development of the two data sets through the Mesolithic. The number of shoreline dated sites undergoes some process of overall decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period. The precise nature of this discrepancy will necessitate further substantiation and the method of shoreline dating itself still requires further testing and assessment. Having noted this, we still tentatively suggest that while the number of shoreline dated sites is not devoid of influence from a demographic signal, this measure could be more heavily influenced by mobility patterns. Conversely, we also suggest that the lacking radiocarbon data from the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data more directly reflects population dynamics.
keywords: | 
  Mesolithic; Shoreline dates; Radiocarbon dates; Fennoscandia
---
Keywords: `r rmarkdown::metadata$keywords`

@@

Previously, the frequency of shoreline dated sites has been evaluated by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g. @fossum2020; @solheim2018; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021]. However, the point-based estimates do not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the north-east to Arendal municipality in the south-west (see map in Fig. \@ref(fig:map)). The result of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.

@@

Summing the probability of the shoreline dated sites and the model-fitting procedures followed the same structure as that for radiocarbon dates and was partly based on re-purposed programming code from *rcarbon* and *ADMUR*. However, idiosyncrasies in the dating method did necessitate some adjustments. To illustrate this, the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Fig. \@ref(fig:shoredate) and outlined below.

Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Fig. \@ref(fig:shoredate)B). These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory of relative sea-level (RSL) change has been the same [see @svendsen1987, and map in Fig. \@ref(fig:shoredate)A]. The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated. This is done by inverse distance weighting [e.g. @conolly2020], where the RSL change is interpolated to the location of an archaeological site based on its distance from the isobases of the geological displacement curves.

@@

With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline. Fig. \@ref(fig:shoredate)C indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve. As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date. However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken. Furthermore, the common practice of adding a constant error or binning point-estimates of shoreline dates, as mentioned above, does not consider that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.

@@

The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833. This is incorporated in the procedure for shoreline dating used for this study (Fig. \@ref(fig:shoredate)D). When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure thus gives the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore be commutative [see @weninger2015, 545], each shoreline date is normalised to sum to unity before being summed. To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.

@@

The implications of different approaches to shoreline dating and summing the results are illustrated in Fig. \@ref(fig:sumshore). In Fig. \@ref(fig:sumshore)A the sites to be shoreline dated are plotted according to their elevation. Due to the variable rates of shoreline displacement within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly inform us about their temporal distribution. For example, as the trajectory for sea-level regression is more pronounced towards the north-east, the same elevation at a location to the south-west implies a younger date compared to one situated further north-east.

When the frequency distribution of shoreline dates has been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking the mean of the TPQ dates for each site (as illustrated in Fig. \@ref(fig:shoredate)C). As outlined above,  these point-estimates are then typically aggregated in bins of 200 or 500 years. This approach is illustrated in Fig. \@ref(fig:sumshore)B. The SSPD resulting from using the probabilistic method of shoreline dating, given in Fig. \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is indicated by the shifts in the positions of peaks and troughs in the distribution after c. 8000 BCE.

@@

In the same way as with an RSPD, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above. Furthermore, similar to how characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of RSL change effectively functions as the calibration curve for each site. As the shoreline displacement curves are interpolated to the sites based on their location along a south-west/north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on exactly the same isobase. With analogy to the radiocarbon methodology, this would be equivalent to each radiocarbon date being associated with a unique calibration curve. As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2 km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites. These intervals were assigned a weight based on how the density of observed sites is distributed among them (Fig. \@ref(fig:incpolys)).

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline-dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
The results of the Monte Carlo simulations are given in Fig. \@ref(fig:spdmc), displaying the deviations from the simulation envelopes and the corresponding global p-values. The comparison of the models based on BIC scores is given in Fig. \@ref(fig:bic).

For the RSPD, the logistic model achieved a p-value of 0.072 and could therefore not be rejected (Fig. \@ref(fig:spdmc)E). As is also indicated by the BIC values of the compared models, the logistic model represents the best explanation of the data, and the failure to reject the model means that the data could be expected under this logistic development. This is characterised by a rapid growth from the start of the RSPD, around 8500 BCE, which plateaus at around 7500 BCE and then remains stable for the remainder of the period.

All the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline-dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Fig. \@ref(fig:spdmc)A). Some significant deviations occur with the model over-predicting the frequency of sites in the period from c. 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE, lasting until around 8500 BCE. The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation between c. 6800 and 5800 BCE. Finally, a positive deviation occurs again around 5300--5000 BCE. While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [@roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded in the discussion to follow.

@@

Another relevant point to mention here is that the RSL change is effectively built into the Monte Carlo simulations, much in the same way as the Monte Carlo simulation method used with radiocarbon dates account for fluctuations in the calibration curve [e.g. @crema2022]. The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Fig. \@ref(fig:incpolys)A), the corresponding local trajectories of RSL change (Fig. \@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma function employed for the dating procedure (Fig. \@ref(fig:shoredate)D). Beyond uncertainties to do with the dating method itself, and by extension also the generation of the Monte Carlo envelope, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models. An increased movement instantiated by increased rates of RSL change could result in the transition to a settlement pattern involving more frequent moves than that which could be expected under a constant adjustment to the shoreline. If this is the case, the proxy would still reflect the degree of occupational duration, but this could be a response to the rate of RSL change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments. Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the substantive implications of the fluctuations in this proxy.

@@

# Acknowledgements{-}
We would like to thank David Wright for commenting on an earlier draft of this paper. We would also like to thank the reviewers for their constructive feedback and PCI archaeology for facilitating their evaluation of our manuscript.
